---
title: "STA457 Project"
author: "Xing Yu Wang"
date: "2025-03-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
# install.packages("forecast")
# install.packages("astsa")

library(dplyr)
library(tidyverse)
library(readr)
library(lubridate)
library(forecast)
library(astsa)
library(tseries)
library(mgcv)
```


# 1. EDA
```{r}
price = read.csv("./Daily Prices_ICCO.csv")
weather = read.csv("./Ghana_data.csv")
```

## 1.1 Clean Data
```{r}
weather <- weather |> dplyr::select(DATE, TAVG)
exchangerate <- USD_GHS_Historical_Data |> dplyr::select(Date, Price)
```

```{r}
colnames(price)[colnames(price) == 'ICCO.daily.price..US..tonne.'] <- 'Daily_Price'
colnames(weather)[colnames(weather) == 'DATE'] <- 'Date'
colnames(weather)[colnames(weather) == 'TAVG'] <- 'Avg_Temp'
colnames(exchangerate)[colnames(exchangerate) == 'Price'] <- 'exchange_rate'
```

## 1.2 Check duplicated values
```{r}
price |> group_by(Date) |> filter(n() > 1) |> ungroup()
```

```{r}
price <- price |> filter(!(Date == "31/01/2024" & Daily_Price == "10,888.05"))
price <- price |> filter(!(Date == "30/01/2024" & Daily_Price == "10,676.42"))
price <- distinct(price)
```

## 1.3 Convert to Time Series Data
### 1.3.1 price Dataset
```{r}
price$Date <- as.Date(price$Date, format="%d/%m/%Y")
price$Daily_Price <- as.numeric(gsub(",", "", price$Daily_Price))
price_month <- price |> mutate(Time = floor_date(Date, "month")) |> group_by(Time) |> 
  summarise(month_Price = mean(Daily_Price, na.rm = TRUE)) |> ungroup()
```

```{r}
summary(price)
```

```{r}
price_ts <- ts(price_month$month_Price, start = c(1994, 11), end = c(2024, 11), frequency = 12)
```

```{r}
plot(price_ts, main="Monthly Price Time Series", ylab="Price", xlab="Time")
```

```{r}
acf2(price_ts, 50)
```

```{r}
ndiffs(price_ts)
```

```{r}
price_month$price_log <- log(price_month$month_Price)
adf.test(price_month$price_log)
```

```{r}
acf2(price_month$price_log, 50)
```

Hence, we want to difference the price data.

```{r}
diff_log_price = diff(price_month$price_log)
ts.plot(diff_log_price, main = "Log Differenced Price Data", ylab = "Log Differenced Price")
```

```{r}
adf.test(diff_log_price)
```

```{r}
acf2(diff_log_price, 50)
```

### 1.3.2 ghana Dataset
```{r}
weather$Date <- as.Date(weather$Date)
weather$Avg_Temp <- as.numeric(gsub("", "", weather$Avg_Temp))
weather_month <- weather |> mutate(Time = floor_date(Date, "month")) |> group_by(Time) |> 
  summarise(Avg_Temp = mean(Avg_Temp, na.rm = TRUE)) |> ungroup()
```

```{r}
summary(weather_month)
```

```{r}
weather_ts <- ts(weather_month$Avg_Temp, start = c(1994, 11), end = c(2024, 11), frequency = 12)
```

```{r}
ts.plot(weather_ts, main="Monthly Average Temperature Time Series", ylab="Temperature", xlab="Time")
```

### 1.3.3 exchange Data
```{r}
exchangerate$Date <- as.Date(exchangerate$Date)
exchangerate$exchange_rate <- as.numeric(gsub("", "", exchangerate$exchange_rate))
rate_month <- exchangerate |> mutate(Time = floor_date(Date, "month")) |> group_by(Time) |> 
  summarise(exchange_rate = mean(exchange_rate, na.rm = TRUE)) |> ungroup()
```

```{r}
summary(exchangerate)
```

```{r}
rate_ts <- ts(rate_month$exchange_rate, start = c(1994, 11), end = c(2024, 11), frequency = 12)
```

```{r}
ts.plot(rate_ts, main="Monthly Average Exchange Rate Time Series", ylab="Exchange Rate (USD/GHS)", xlab="Time")
```

```{r}
par(mfrow=c(3,1), mar = c(3, 4, 2, 2))
# price
plot(price_ts, main="Monthly Price Time Series", ylab="Price", xlab="Time")
#temperature
ts.plot(weather_ts, main="Monthly Average Temperature Time Series", ylab="Temperature", xlab="Time")
# exchange rate
ts.plot(rate_ts, main="Monthly Average Exchange Rate Time Series", ylab="Exchange Rate (USD/GHS)", xlab="Time")
```

## 1.4 Time series plots for data
```{r}
ggtsdisplay(price_ts, main="Time series plot for Monthly World Cocoa Price")
```


```{r}
ggtsdisplay(weather_ts, main="Time series plot for Monthly Average Temperature")
```

```{r}
ggtsdisplay(rate_ts, main="Time series plot for Monthly Average Exchange Rate(USD/GHS)")
```

```{r}
par(mfrow=c(3,1), mar = c(3, 4, 2, 2))
# price
plot(price_ts, main="Monthly Price Time Series", ylab="Price", xlab="Time")
#temperature
ts.plot(weather_ts, main="Monthly Average Temperature Time Series", ylab="Temperature", xlab="Time")
# exchange rate
ts.plot(rate_ts, main="Monthly Average Exchange Rate Time Series", ylab="Exchange Rate (USD/GHS)", xlab="Time")
```
## 1.5 Combine and Split data
```{r}
data <- price_month |> left_join(weather_month, by = "Time") |> left_join(rate_month, by = "Time")
data <- data |> mutate(log_price = log(month_Price), diff_log_price = 
                         c(NA, diff(price_month$price_log))) |> drop_na()
data <- data |> dplyr::select(Time, Avg_Temp, exchange_rate, diff_log_price, log_price, month_Price)

data$Time <- as.Date(data$Time)
```

```{r}
set.seed(889)

sample_size <- floor(0.7 * nrow(data))
train_indices <- sample(seq_len(nrow(data)), size = sample_size)
trainSet <- data[train_indices, ]
testSet <- data[-train_indices, ]
```

```{r}
data_train_ts <- ts(trainSet$diff_log_price, frequency = 12)
```

```{r}
diff_price_ts <- ts(data$diff_log_price, frequency = 12)
diff_price_ts_components <- decompose(diff_price_ts)
plot(diff_price_ts_components)
```


# 2. Method
## 2.1 ETS Model
ETS is a purely univariate model and cannot directly handle external regressors.

### 2.1.1 Fit Model
```{r}
ets_model <- ets(data_train_ts)
ets_zmodel <- ets(data_train_ts, model = "ZZZ") # Automatically selects best model
summary(ets_model)
summary(ets_zmodel)
```
ETS(A,N,N) is the best model.

### 2.1.2 Forecasting and Plotting
```{r}
# Plot using log differenced price
data_test_ts <- ts(testSet$diff_log_price, start = end(data_train_ts) + c(0,1), 
                   frequency = 12)

h <- nrow(testSet)
forecast_ets <- forecast(ets_model, h = h)

autoplot(forecast_ets) + autolayer(data_test_ts, series = "Actual", color = "red")
```
The red line is the observed actual values. The forecasted values are the central blue line within the blue shaded prediction intervals.

```{r}
last_log_price <- tail(trainSet$log_price , 1)
forecasted_log_price <- cumsum(forecast_ets$mean) + last_log_price

# Convert back to actual price
forecasted_price <- exp(forecasted_log_price)

actual_price <- exp(testSet$log_price)
```

```{r}
data_test_ts <- ts(testSet$diff_log_price, start = end(data_train_ts) + c(0,1), 
                   frequency = 12)

forecast_ets_ts <- ts(forecasted_price, start = start(data_test_ts), frequency = 12)
actual_ets_ts <- ts(actual_price, start = start(data_test_ts), frequency = 12)
```

```{r}
# Plot using actual price
autoplot(forecast_ets_ts, series = "Predicted") + 
  autolayer(actual_ets_ts, series = "Actual", color = "blue") +
  ggtitle("Forecast vs Actual Prices") +
  ylab("Price") +
  xlab("Time") +
  theme_minimal()
```

## 2.2 ARIMAX Model

Recall that in Section 1.3.1, we have tested the acf and adf.test, and determined that we would be using the differenced price data. To fit the trainset, we evaluate p and q for ARIMA model.

```{r}
adf.test(trainSet$diff_log_price)
```


```{r}
acf2(trainSet$diff_log_price)
```

### 2.2.1 Fit ARIMAX Model
```{r}
xreg_matrix <- cbind(trainSet$Avg_Temp, trainSet$exchange_rate)

# Assign column names to xreg
colnames(xreg_matrix) <- c("Avg_Temp", "exchange_rate")

# Fit the SARIMA model with the named xreg
arimax_model <- Arima(trainSet$diff_log_price, order=c(1,0,1), xreg = xreg_matrix)

# Summary of the model
summary(arimax_model)
```

```{r}
checkresiduals(arimax_model)
```

Fail to reject $H_0$, hence residuals of this plot do not show significant autocorrelation.

### 2.2.2 Forecasting With ARIMAX Model

```{r}
# Randomly sample 5 rows from testSet for exogenous variables
set.seed(123)  # For reproducibility
random_rows <- sample(nrow(testSet), 15)

# Create future xreg from randomly sampled rows
forecast_arimax_xreg <- cbind(testSet$Avg_Temp, testSet$exchange_rate)

# Ensure column names match the original xreg
colnames(forecast_arimax_xreg) <- c("Avg_Temp", "exchange_rate")

# Forecast using randomly selected xreg
forecast_arimax <- predict(arimax_model, newdata=testSet, newxreg=forecast_arimax_xreg)
```


```{r}
# last_log_price <- tail(trainSet$log_price , 1)
forecasted_log_price_arimax <- cumsum(forecast_arimax$mean) + last_log_price

# Convert back to actual price
forecasted_price_arimax <- exp(forecasted_log_price_arimax)
actual_price_arimax <- testSet$month_Price

forecast_arimax_ts <- ts(forecasted_price_arimax, start = start(data_test_ts), frequency = 12)
actual_arimax_ts <- ts(actual_price_arimax, start = start(data_test_ts), frequency = 12)
```

```{r}
# Plot using actual price
autoplot(forecast_arimax_ts, series = "Predicted") + 
  autolayer(actual_arimax_ts, series = "Actual", color = "blue") +
  ggtitle("Forecast vs Actual Prices") +
  ylab("Price") +
  xlab("Time") +
  theme_minimal()
```


## 2.3 VAR model
```{r}
library(vars)
library(dplyr)
library(forecast)
```

```{r}
# Prepare the data: price, exchange rate, and average temperature
var_data <- trainSet %>% dplyr::select(where(is.numeric))
VARselect(var_data, lag.max = 10, type = "const")
var_model <- VAR(var_data, p=1, type="const")
summary(var_model)
```
Not a very reliable model for our dataset :).

```{r}
library(forecast)
var_data <- trainSet %>% dplyr::select(where(is.numeric))
xreg_matrix <- as.matrix(cbind(var_data$Avg_Temp, var_data$exchange_rate))

adf.test(var_data$diff_log_price)

armax_model <- auto.arima(var_data$diff_log_price, xreg = xreg_matrix, d=0)
summary(armax_model)
```
not sure if this is the right way to fit ARMAX model, cuz in the book they are actually using an example of VAR under Multivariate ARMAX section.

```{r}
plot(var_data$diff_log_price, type="l")
```


```{r}
summary(xreg_matrix)
cor(xreg_matrix)
```

```{r}
library(TSA)
library(forecast)
air.m1<-arimax(log(airmiles),order=c(0,0,1),
              xtransf=data.frame(I911=1*(seq(airmiles)==69)),
              transfer=list(c(1,0))
              )
```

## 2.5 GAM Model
### 2.5.1 Fit Model
#### 2.5.1.1 Basic Model
```{r}
# Uses simple smoothing splines for variables
trainSet$Time_num <- as.numeric(trainSet$Time)

gam_basic <- gam(diff_log_price ~ s(Time_num) + s(Avg_Temp) + s(exchange_rate), 
                 data = trainSet, method = "REML")
```

```{r}
gam.check(gam_basic)
```

#### 2.5.1.2 Complex Model
```{r}
# Fit GAM model with smooth trend, random effects, and seasonality
trainSet$dateInt = as.integer(trainSet$Time)
trainSet$monthFac <- factor(format(trainSet$Time, "%m"))
gam_complex <- gam(diff_log_price ~ s(dateInt, k = 100) + s(monthFac, bs = "re") + 
                     sinpi(dateInt / 182.625) + cospi(dateInt / 182.625) + 
                     sinpi(dateInt / 91.3125) + cospi(dateInt / 91.3125) + 
                     s(Avg_Temp) + s(exchange_rate), data = trainSet, 
                   method = "REML", optimizer = "efs")
```

```{r}
summary(gam_basic)
summary(gam_complex)
```

```{r}
plot(gam_complex, pages = 1, shade = TRUE)
```
```{r}
gam.check(gam_complex) 
```






