---
title: "STA457 Project"
author: "Xing Yu Wang"
date: "2025-03-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
# install.packages("forecast")
# install.packages("astsa")

library(dplyr)
library(tidyverse)
library(readr)
library(lubridate)
library(forecast)
library(astsa)
library(tseries)
```


# 1. EDA
```{r}
price = read.csv("./Daily Prices_ICCO.csv")
weather = read.csv("./Ghana_data.csv")
```

## 1.1 Clean Data
```{r}
weather <- weather |> select(DATE, TAVG)
exchangerate <- USD_GHS_Historical_Data |> select(Date, Price)
```

```{r}
colnames(price)[colnames(price) == 'ICCO.daily.price..US..tonne.'] <- 'Daily_Price'
colnames(weather)[colnames(weather) == 'DATE'] <- 'Date'
colnames(weather)[colnames(weather) == 'TAVG'] <- 'Avg_Temp'
colnames(exchangerate)[colnames(exchangerate) == 'Price'] <- 'exchange_rate'
```

## 1.2 Check duplicated values
```{r}
price |> group_by(Date) |> filter(n() > 1) |> ungroup()
```

```{r}
price <- price |> filter(!(Date == "31/01/2024" & Daily_Price == "10,888.05"))
price <- price |> filter(!(Date == "30/01/2024" & Daily_Price == "10,676.42"))
price <- distinct(price)
```

## 1.3 Convert to Time Series Data
### 1.3.1 price Dataset
```{r}
price$Date <- as.Date(price$Date, format="%d/%m/%Y")
price$Daily_Price <- as.numeric(gsub(",", "", price$Daily_Price))
price_month <- price |> mutate(Time = floor_date(Date, "month")) |> group_by(Time) |> 
  summarise(Daily_Price = mean(Daily_Price, na.rm = TRUE)) |> ungroup()
```

```{r}
price_ts <- ts(price_month$Daily_Price, start = c(1994, 11), end = c(2024, 11), frequency = 12)
```

```{r}
plot(price_ts, main="Monthly Price Time Series", ylab="Price", xlab="Time")
```

```{r}
acf2(price_ts, 50)
```

```{r}
ndiffs(price_ts)
```

```{r}
price_month$price_log <- log(price_month$Daily_Price)
diff_log_price = diff(price_month$price_log)
```

```{r}
ts.plot(diff_log_price, main = "Log Differenced Price Data", ylab = "Log Differenced Price")
```
```{r}
adf.test(diff_log_price)
```
```{r}
acf2(diff_log_price, 50)
```



### 1.3.2 ghana Dataset
```{r}
weather$Date <- as.Date(weather$Date)
weather$Avg_Temp <- as.numeric(gsub("", "", weather$Avg_Temp))
weather_month <- weather |> mutate(Time = floor_date(Date, "month")) |> group_by(Time) |> 
  summarise(Avg_Temp = mean(Avg_Temp, na.rm = TRUE)) |> ungroup()
```

```{r}
weather_ts <- ts(weather_month$Avg_Temp, start = c(1994, 11), end = c(2024, 11), frequency = 12)
```

```{r}
ts.plot(weather_ts, main="Monthly Average Temperature Time Series", ylab="Temperature", xlab="Time")
```

### 1.3.3 exchange Data
```{r}
exchangerate$Date <- as.Date(exchangerate$Date)
exchangerate$exchange_rate <- as.numeric(gsub("", "", exchangerate$exchange_rate))
rate_month <- exchangerate |> mutate(Time = floor_date(Date, "month")) |> group_by(Time) |> 
  summarise(exchange_rate = mean(exchange_rate, na.rm = TRUE)) |> ungroup()
```

```{r}
rate_ts <- ts(rate_month$exchange_rate, start = c(1994, 11), end = c(2024, 11), frequency = 12)
```

```{r}
ts.plot(rate_ts, main="Monthly Average Exchange Rate Time Series", ylab="Exchange Rate (USD/GHS)", xlab="Time")
```

## 1.4 Combine and Split data
```{r}
ggtsdisplay(price_ts, main="Time series plot for Monthly World Cocoa Price")
```

```{r}
ggtsdisplay(weather_ts, main="Time series plot for Monthly Average Temperature")
```

```{r}
ggtsdisplay(rate_ts, main="Time series plot for Monthly Average Exchange Rate(USD/GHS)")
```


```{r}
data <- price_month |> left_join(weather_month, by = "Time") |> left_join(rate_month, by = "Time")
data <- data |> mutate(log_price = log(Daily_Price), diff_log_price = 
                         c(NA, diff(price_month$price_log))) |> drop_na()
data <- data |> select(Time, Avg_Temp, exchange_rate, diff_log_price)

data$Time <- as.Date(data$Time)
```

```{r}
set.seed(889)

sample_size <- floor(0.7 * nrow(data))
train_indices <- sample(seq_len(nrow(data)), size = sample_size)
trainSet <- data[train_indices, ]
testSet <- data[-train_indices, ]
```

```{r}
data_train_ts <- ts(trainSet$diff_log_price, frequency = 12)
```

# 2. Method
## 2.1 ETS Model
ETS is a purely univariate model and cannot directly handle external regressors.

### 2.1.1 Fit Model
```{r}
ets_model <- ets(data_train_ts)
ets_zmodel <- ets(data_train_ts, model = "ZZZ") # Automatically selects best model
summary(ets_model)
summary(ets_zmodel)
```
ETS(A,N,N) is the best model.

### 2.1.2 Forecasting
```{r}
data_test_ts <- ts(testSet$diff_log_price, start = end(data_train_ts) + c(0,1), 
                   frequency = 12)

h <- nrow(testSet)
forecast_ets <- forecast(ets_model, h = h)

autoplot(forecast_ets) + autolayer(data_test_ts, series = "Actual", color = "red")
```
The red line is the observed actual values. The forecasted values are the central blue line within the blue shaded prediction intervals.

## 2.2 ARIMAX Model

```{r}
adf.test(price_month$price_log)
```

```{r}
acf2(price_month$price_log)
```


```{r}
acf2(trainSet$diff_log_price, 50)
```


## 2.3 VAR model
```{r}
# install.packages("vars")
library(vars)
library(dplyr)
library(forecast)
```

```{r}
# Prepare the data: price, exchange rate, and average temperature
var_data <- trainSet %>% dplyr::select(where(is.numeric))
VARselect(var_data, lag.max = 10, type = "const")
var_model <- VAR(var_data, p=1, type="const")
summary(var_model)
```
Not a very reliable model for our dataset :).


```{r}
library(forecast)
var_data <- trainSet %>% dplyr::select(where(is.numeric))
xreg_matrix <- as.matrix(cbind(var_data$Avg_Temp, var_data$exchange_rate))

adf.test(var_data$diff_log_price)

armax_model <- auto.arima(var_data$diff_log_price, xreg = xreg_matrix, d=0)
summary(armax_model)
```
not sure if this is the right way to fit ARMAX model, cuz in the book they are actually using an example of VAR under Multivariate ARMAX section.

```{r}
plot(var_data$diff_log_price, type="l")
```


```{r}
summary(xreg_matrix)
cor(xreg_matrix)
```

```{r}
library(TSA)
library(forecast)
air.m1<-arimax(log(airmiles),order=c(0,0,1),
              xtransf=data.frame(I911=1*(seq(airmiles)==69)),
              transfer=list(c(1,0))
              )
```



