---
title: "STA457 Project"
author: "Xingyu Wang, Xinyue Tao"
date: "2025-03-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", out.width = "75%")
```

```{r, message=FALSE}
library(dplyr)
library(tidyverse)
library(readr)
library(lubridate)
library(forecast)
library(astsa)
library(tseries)
library(mgcv)
library(Metrics)
library(ggplot2)
library(xgboost)
library(Matrix)
library(caret)
library(rugarch)
library(tibble)
library(xts)
library(gridExtra)
```

# 1. EDA
```{r}
price = read.csv("./Daily Prices_ICCO.csv")
weather = read.csv("./Ghana_data.csv")
# USD_GHS_Historical_Data <- read_csv("~/sta457/STA457_Project/Project/USD_GHS Historical Data.csv")
USD_GHS_Historical_Data = read.csv("./USD_GHS Historical Data.csv")
```

## 1.1 Clean Data
```{r}
weather <- weather |> dplyr::select(DATE, TAVG)
exchangerate <- USD_GHS_Historical_Data |> dplyr::select(Date, Price)
```

```{r}
colnames(price)[colnames(price) == 'ICCO.daily.price..US..tonne.'] <- 'Daily_Price'
colnames(weather)[colnames(weather) == 'DATE'] <- 'Date'
colnames(weather)[colnames(weather) == 'TAVG'] <- 'Avg_Temp'
colnames(exchangerate)[colnames(exchangerate) == 'Price'] <- 'exchange_rate'
```

## 1.2 Check duplicated values
```{r}
price |> group_by(Date) |> filter(n() > 1) |> ungroup()
```

```{r}
price <- price |> filter(!(Date == "31/01/2024" & Daily_Price == "10,888.05"))
price <- price |> filter(!(Date == "30/01/2024" & Daily_Price == "10,676.42"))
price <- distinct(price)
```

## 1.3 Convert to Time Series Data
### 1.3.1 price Dataset
```{r}
price$Date <- as.Date(price$Date, format="%d/%m/%Y")
price$Daily_Price <- as.numeric(gsub(",", "", price$Daily_Price))
price_month <- price |> mutate(Time = floor_date(Date, "month")) |> group_by(Time) |> 
  summarise(month_Price = mean(Daily_Price, na.rm = TRUE)) |> ungroup()
```

```{r}
summary(price)
```

```{r}
price_ts <- ts(price_month$month_Price, start = c(1994, 11), end = c(2024, 11), frequency = 12)
```

```{r}
plot(price_ts, main="Monthly Price Time Series", ylab="Price", xlab="Time")
```

```{r}
acf2(price_ts, 50)
```

```{r}
ndiffs(price_ts)
```

```{r}
price_month$price_log <- log(price_month$month_Price)
adf.test(price_month$price_log)
```

```{r}
acf2(price_month$price_log, 50)
```

Hence, we want to difference the price data.

```{r}
diff_log_price = diff(price_month$price_log)
ts.plot(diff_log_price, main = "Log Differenced Price Data", ylab = "Log Differenced Price")
```

```{r}
adf.test(diff_log_price)
```

```{r}
acf2(diff_log_price, 50)
```

### 1.3.2 ghana Dataset
```{r}
weather$Date <- as.Date(weather$Date)
weather$Avg_Temp <- as.numeric(gsub("", "", weather$Avg_Temp))
weather_month <- weather |> mutate(Time = floor_date(Date, "month")) |> group_by(Time) |> 
  summarise(Avg_Temp = mean(Avg_Temp, na.rm = TRUE)) |> ungroup()
```

```{r}
summary(weather_month)
```

```{r}
weather_ts <- ts(weather_month$Avg_Temp, start = c(1994, 11), end = c(2024, 11), frequency = 12)
```

```{r}
ts.plot(weather_ts, main="Monthly Average Temperature Time Series", ylab="Temperature", xlab="Time")
```

### 1.3.3 exchange Data
```{r}
exchangerate$Date <- as.Date(exchangerate$Date)
exchangerate$exchange_rate <- as.numeric(gsub("", "", exchangerate$exchange_rate))
rate_month <- exchangerate |> mutate(Time = floor_date(Date, "month")) |> group_by(Time) |> 
  summarise(exchange_rate = mean(exchange_rate, na.rm = TRUE)) |> ungroup()
```

```{r}
summary(exchangerate)
```

```{r}
rate_ts <- ts(rate_month$exchange_rate, start = c(1994, 11), end = c(2024, 11), frequency = 12)
```

```{r}
ts.plot(rate_ts, main="Monthly Average Exchange Rate Time Series", ylab="Exchange Rate (USD/GHS)", xlab="Time")
```

```{r}
par(mfrow=c(3,1), mar = c(3, 4, 2, 2))
# price
plot(price_ts, main="Monthly Price Time Series", ylab="Price", xlab="Time")
#temperature
ts.plot(weather_ts, main="Monthly Average Temperature Time Series", ylab="Temperature", xlab="Time")
# exchange rate
ts.plot(rate_ts, main="Monthly Average Exchange Rate Time Series", ylab="Exchange Rate (USD/GHS)", xlab="Time")
```

## 1.4 Time series plots for data
```{r}
ggtsdisplay(price_ts, main="Time series plot for Monthly World Cocoa Price")
```

```{r}
ggtsdisplay(weather_ts, main="Time series plot for Monthly Average Temperature")
```

```{r}
ggtsdisplay(rate_ts, main="Time series plot for Monthly Average Exchange Rate(USD/GHS)")
```

```{r}
par(mfrow=c(3,1), mar = c(3, 4, 2, 2))
# price
plot(price_ts, main="Monthly Price Time Series", ylab="Price", xlab="Time")
#temperature
ts.plot(weather_ts, main="Monthly Average Temperature Time Series", ylab="Temperature", xlab="Time")
# exchange rate
ts.plot(rate_ts, main="Monthly Average Exchange Rate Time Series", ylab="Exchange Rate (USD/GHS)", xlab="Time")
```


## 1.5 Combine Datasets
```{r}
data <- price_month |> left_join(weather_month, by = "Time") |> left_join(rate_month, by = "Time")
data <- data |> mutate(log_price = log(month_Price), diff_log_price = 
                         c(NA, diff(price_month$price_log))) |> drop_na()
data <- data |> dplyr::select(Time, Avg_Temp, exchange_rate, diff_log_price, log_price, month_Price)

data$Time <- as.Date(data$Time)
```

```{r}
plot(data$Avg_Temp, data$month_Price, xlab = "Monthly Price", ylab = "Average Temperature", 
     main = "Daily Price vs. Avg Temperature")
pairs(data[, c("month_Price", "Avg_Temp", "exchange_rate")])
```


## 1.6 Stationary Check
```{r}
adf.test(data$Avg_Temp)
adf.test(data$exchange_rate)
adf.test(log(data$exchange_rate))
adf.test(diff(log(data$exchange_rate)))
```

Since monthly average temperature is already stationary, we would do take the differenced and log-transformed exchange rate as our exogenous factors.

```{r}
acf2(data$exchange_rate)
acf2(log(data$exchange_rate))
acf2(diff(log(data$exchange_rate)))
```

ACF shows similar trend, where only differenced log-transformed exchange rate is stationary. Hence, this differenced and log-transformed exchange rate will be used as one of the external(exogenous) regressors in ARIMAX and GARCHX.

```{r}
adf.test(data$month_Price)
adf.test(data$log_price)
adf.test(data$diff_log_price)
```

Since only the diff_log_price is stationary, we choose differenced monthly log price when fitting the model.

```{r}
diff_price_ts <- ts(data$diff_log_price, frequency = 12)
autoplot(decompose(diff_price_ts, type="additive")) +
  ggtitle("Decomposition of Differenced Log Price Time Series") +
  theme_minimal()
```


## 1.7 Split data
```{r}
data <- data[order(data$Time), ]
cutoff <- floor(0.7 * nrow(data))
trainSet <- data[1:cutoff, ]
testSet <- data[(cutoff+1):nrow(data), ]
```

# 2. Method
## 2.1 ETS Model
ETS is a purely univariate model and cannot directly handle external regressors.
```{r}
data_train_ts <- ts(trainSet$diff_log_price, frequency = 12)
```

### 2.1.1 Fit Model
```{r}
ets_model <- ets(data_train_ts, model = "ANA")
ets_zmodel <- ets(data_train_ts, model = "ZZZ") # Automatically selects best model
summary(ets_model)
summary(ets_zmodel)
```

```{r}
plot(ets_model)
plot(ets_zmodel)
```

### 2.1.2 Forecasting and Plotting
```{r}
# Plot using log differenced price
data_test_ts <- ts(testSet$diff_log_price, start = end(data_train_ts) + c(0,1), 
                   frequency = 12)

h <- nrow(testSet)
forecast_ets <- forecast(ets_model, h = h)
forecast_zets <- forecast(ets_zmodel, h = h)

autoplot(forecast_ets) + autolayer(data_test_ts, series = "Actual", color = "red")
autoplot(forecast_zets) + autolayer(data_test_ts, series = "Actual", color = "red")
```
The red line is the observed actual values. The forecasted values are the central blue line within the blue shaded prediction intervals.

```{r}
last_log_price <- tail(trainSet$log_price, 1)

# Convert back to actual price
forecasted_price <- exp(cumsum(forecast_ets$mean) + last_log_price)
forecasted_zprice <- exp(cumsum(forecast_zets$mean) + last_log_price)

actual_price <- exp(testSet$log_price)
```

```{r}
data_test_ts <- ts(testSet$diff_log_price, start = end(data_train_ts) + c(0,1), 
                   frequency = 12)

forecast_ets_ts <- ts(forecasted_price, start = start(data_test_ts), frequency = 12)
forecast_zets_ts <- ts(forecasted_zprice, start = start(data_test_ts), frequency = 12)
actual_ets_ts <- ts(actual_price, start = start(data_test_ts), frequency = 12)
```

```{r}
# ANA
p1 <- autoplot(forecast_ets_ts, series = "Prediction") + 
  autolayer(actual_ets_ts, series = "Actual") +
  ggtitle("ETS(ANA): Predicted vs Actual Prices") +
  ylab("Price") +
  xlab("Time") +
  theme_minimal()+
  scale_color_manual(values = c(
    "Actual" = "black",
    "Prediction" = "red"))
p1
```

```{r}
# ANN
autoplot(forecast_zets_ts, series = "Predicted") + 
  autolayer(actual_ets_ts, series = "Actual", color = "blue") +
  ggtitle("Forecast vs Actual Prices") +
  ylab("Price") +
  xlab("Time") +
  theme_minimal()
```

```{r}
checkresiduals(ets_model)
checkresiduals(ets_zmodel)
```

```{r}
# RMSE
sqrt(mean((actual_price - forecasted_price)^2))
sqrt(mean((actual_price - forecasted_zprice)^2))

#MAE
mean(abs(actual_ets_ts - forecast_ets_ts))
mean(abs(actual_ets_ts - forecast_zets_ts))

# MAPE
mean(abs((actual_ets_ts - forecast_ets_ts) / actual_ets_ts)) * 100
mean(abs((actual_ets_ts - forecast_zets_ts) / actual_ets_ts)) * 100
```

## 2.2 ARIMAX Model

Recall that in Section 1.3.1, we have tested the acf and adf.test, and determined that we would be using the differenced price data before we fit the model.

```{r}
adf.test(diff(trainSet$month_Price))
```

```{r}
adf.test(trainSet$log_price)
```

```{r}
acf2(trainSet$diff_log_price)
```

```{r}
adf.test(data$Avg_Temp)
```


```{r}
adf.test(diff(log(data$exchange_rate)))
```


### 2.2.1 Fit ARIMAX Model

```{r}
dl.rate.train <- c(0, diff(log(trainSet$exchange_rate)))
xreg_matrix <- cbind(trainSet$Avg_Temp, dl.rate.train)
colnames(xreg_matrix) <- c("Avg_Temp", "dl_exchange_rate")

p <- 0:10
q <- 0:10
aic.arimax <- matrix(0, length(p), length(q)) 
for (i in 1:length(p)) {
  for (j in 1:length(q)) {
      modij = Arima(trainSet$diff_log_price, order = c(p[i], 0, q[j]), 
                    method = "ML", xreg=xreg_matrix)
      aic.arimax[i, j] = AIC(modij)
  }
}

aic.arimax.min_index <- which(aic.arimax == min(aic.arimax), arr.ind = TRUE)
ariamx.p <- p[aic.arimax.min_index[1]]
ariamx.q <- q[aic.arimax.min_index[2]]
sprintf("Selected order for ARMA: p = %d, q = %d", ariamx.p, ariamx.q)
```

```{r}
model.arimax <- Arima(trainSet$diff_log_price, order=c(ariamx.p,0,ariamx.q), xreg = xreg_matrix)
summary(model.arimax)
```

```{r}
checkresiduals(model.arimax)
```

```{r}
qqnorm(model.arimax$residuals)
qqline(model.arimax$residuals, col="red")
```
```{r}
adf.test(model.arimax$residuals)
```

- ADF Test on ARIMAX Model Residuals: Failed to reject H_0, indicating that the residuals do not exhibit significant autocorrelation.
- Histogram and QQ-Plot of Residuals: Residuals align well with the 45-degree line, suggesting normality.
- ACF of Residuals: Appears random, with all lags within the range of -0.15 to 0.1, indicating no strong autocorrelations.
- Standardized Residuals Plot: No discernible trend observed, further supporting the model’s adequacy.
- Ljung-Box Test (Residuals from ARIMA(2,0,3) model):
  - Q^* = 7.1814, df = 5, p-value = 0.2075
  - Model degrees of freedom: 5, Total lags used: 10
Conclusion: The ARIMAX model effectively captures the trend of the training dataset.

### 2.2.2 Forecasting With ARIMAX Model
Next we try to fit this ARIMAX model to forecast on testing set.

```{r}
dl.rate.test <- c(0, diff(log(testSet$exchange_rate)))
forecast.arimax.xreg <- cbind(testSet$Avg_Temp, dl.rate.test)
colnames(forecast.arimax.xreg) <- c("Avg_Temp", "dl_exchange_rate")
forecast.arimax <- forecast(model.arimax, xreg=forecast.arimax.xreg, 
                            h=nrow(testSet))

last_log_price <- tail(trainSet$log_price, 1)
# Convert back to actual price
forecast.arimax.final <- exp(cumsum(forecast.arimax$mean) + last_log_price)
```

```{r}
model.arimax.fitted <- as.numeric(model.arimax$fitted)
model.arimax.fitted.converted <- exp(log(trainSet$month_Price[1]) 
                                     + cumsum(model.arimax.fitted))
rmse(trainSet$month_Price, model.arimax.fitted.converted)
mae(trainSet$month_Price, model.arimax.fitted.converted)
mape(trainSet$month_Price, model.arimax.fitted.converted)
```

```{r}
rmse(testSet$month_Price, forecast.arimax.final)
mae(testSet$month_Price, forecast.arimax.final)
mape(testSet$month_Price, forecast.arimax.final)
```

```{r}
forecast.arimax.df <- tibble(
  Time = testSet$Time,
  Price = forecast.arimax.final
)
test.arimax.df <- tibble(
  Time = testSet$Time,
  Price = testSet$month_Price
)
```

```{r}
p2 <- ggplot() +
  geom_line(data = test.arimax.df, aes(x = Time, y = Price, color = "Actual")) +
  geom_line(data = forecast.arimax.df, aes(x = Time, y = Price, color = "Prediction")) +
  labs(
    title = "ARIMAX(2,0,3): Predicted vs Actual Prices", 
    y = "Price", 
    x = "Time",
    color = "Legend"
  ) +
  theme_minimal() +
  scale_color_manual(values = c(
    "Actual" = "black",
    "Prediction" = "red"))
p2
```
## 2.3 ARMAX-GARCH Model
### 2.3.1 ARMAX-GARCH Parameters
```{r}
# xreg_matrix is the same as arimax
p = 0:3
q = 0:3
## select ARMA order
aic.armax.garch1 <- matrix(0, length(p), length(q)) 
for (i in 1:length(p)) {
  for (j in 1:length(q)) {
      modij = Arima(trainSet$diff_log_price, order = c(p[i], 0, q[j]), 
                    method = "ML", xreg=xreg_matrix)
      aic.armax.garch1[i, j] = AIC(modij)
  }
}
aic.armax.min_index <- which(aic.armax.garch1 == min(aic.armax.garch1), arr.ind = TRUE)
aramx.garch.p <- p[aic.armax.min_index[1]]
aramx.garch.q <- q[aic.armax.min_index[2]]
sprintf("Selected order for ARMA: p = %d, q = %d", aramx.garch.p, aramx.garch.q)
```

This is the same as what we have for ARIMAX. Then we use the similar method, where we fix the armax order, and systematically search for the combination of orders for garch configuration with smallest AIC.

```{r}
m = 1:3
n = 1:3
# dl.rate.train <- c(0, diff(log(trainSet$exchange_rate)))
# xreg_matrix <- cbind(trainSet$Avg_Temp, dl.rate.train)
# colnames(xreg_matrix) <- c("Avg_Temp", "dl_exchange_rate")
## select GARCH order
aic.armax.garch2 <- matrix(0, length(m), length(n)) 
for (i in 1:length(m)) {
  for (j in 1:length(n)) {
      spec = ugarchspec(variance.model=list(model="sGARCH", 
                                            garchOrder=c(m[i],n[j])),
                mean.model=list(armaOrder=c(aramx.garch.p, aramx.garch.q), 
                                include.mean=T, 
                                external.regressors = xreg_matrix),
                distribution.model="std")   
      modij = ugarchfit(spec=spec, data = trainSet$diff_log_price,
                        trace = FALSE)
      aic.armax.garch2[i, j] = infocriteria(modij)[1]
  }
}


aic.garch.min_index <- which(aic.armax.garch2 == min(aic.armax.garch2), arr.ind = TRUE)
aramx.garch.m <- m[aic.garch.min_index[1]]
aramx.garch.n <- n[aic.garch.min_index[2]]
sprintf("Selected order for GARCH: m = %d, n = %d", aramx.garch.m, aramx.garch.n)
```

```{r}
armax.garch.spec.train <- ugarchspec(variance.model=list(model="sGARCH", 
                                            garchOrder=c(aramx.garch.m,aramx.garch.n)),
                mean.model=list(armaOrder=c(aramx.garch.p, aramx.garch.q), 
                                include.mean=T, 
                                external.regressors = xreg_matrix),
                distribution.model="std")   
model.armax.garch <- ugarchfit(armax.garch.spec.train, 
                               data = trainSet$diff_log_price,
                               trace = FALSE)
model.armax.garch
```

```{r}
model.armax.garch@fit$coef
```

```{r}
garch_time_index <- as.POSIXct(trainSet$Time)
residuals_armax_garch_xts <- xts(residuals(model.armax.garch), 
                                 order.by = garch_time_index)
std_resid_armax_garch_xts <- xts(model.armax.garch@fit$z, 
                                 order.by = garch_time_index)

# Residual Analysis
par(mfrow = c(2, 2))

# Residual plots
plot(residuals_armax_garch_xts, main = "Residuals")
plot(std_resid_armax_garch_xts, main = "Standardized Residuals")

# ACF plots
acf(na.omit(as.numeric(residuals(model.armax.garch))), main = "ACF of Residuals")
acf(na.omit(as.numeric(residuals(model.armax.garch)^2)), main = "ACF of Squared Residuals")
```

```{r}
armax.garch.actual.values <- trainSet$diff_log_price
armax.garch.fitted.values <- as.numeric(fitted(model.armax.garch))

armax.garch.fit.df <- tibble(
  Time = trainSet$Time,
  Price = armax.garch.fitted.values
)
armax.garch.train.df <- tibble(
  Time = trainSet$Time,
  Price = trainSet$diff_log_price
)
```

```{r}
ggplot() +
  geom_line(data = armax.garch.fit.df, aes(x = Time, y = Price, color = "Fitted")) +
  geom_line(data = armax.garch.train.df, aes(x = Time, y = Price, color = "Actual")) +
  labs(
    title = "sGARCH(1,1) with ARFIMA(2,0,3) Model: Fitted vs Actual Prices", 
    y = "Price", 
    x = "Time",
    color = "Legend"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("Actual" = "grey", "Fitted" = "red"))
```

### 2.3.2 ARMAX-GARCH Forecast
```{r}
# multi-step forecast
ngarchfore = length(testSet$diff_log_price)
xreg_test_matrix <- cbind(testSet$Avg_Temp, diff(log(testSet$exchange_rate)))
colnames(xreg_test_matrix) <- c("Avg_Temp", "dl_exchange_rate")

fore.garch.dl = ugarchforecast(model.armax.garch,
                               n.ahead = ngarchfore,
                              external.forecasts = list(mreg=xreg_test_matrix))
fore.garch.dl.data <- as.numeric(fore.garch.dl@forecast$seriesFor)
last_log_price <- tail(trainSet$log_price, 1)
forecast.armax.garch.multi <- exp(cumsum(fore.garch.dl.data) + last_log_price)
```

```{r}
library(tibble)
forecast.garch.multi.df <- tibble(
  Time = testSet$Time,
  Price = forecast.armax.garch.multi
)
test.garch.df <- tibble(
  Time = testSet$Time,
  Price = testSet$month_Price
)
```

```{r}
p3 <- ggplot() +
  geom_line(data = test.garch.df, aes(x = Time, y = Price, color = "Actual")) +
  geom_line(data = forecast.garch.multi.df, aes(x = Time, y = Price, color = "Prediction")) +
  labs(
    title = "GARCH(1,1) with ARMAX(2,3): Predicted vs Actual Prices", 
    y = "Price", 
    x = "Time",
    color = "Legend"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("Actual" = "black", "Prediction" = "red"))
p3
```

```{r}
model.armax.garch.fitted <- as.numeric(fitted(model.armax.garch))
model.armax.garch.fitted.converted <- exp(log(trainSet$month_Price[1]) 
                                     + cumsum(model.armax.garch.fitted))
rmse(trainSet$month_Price, model.armax.garch.fitted.converted)
mae(trainSet$month_Price, model.armax.garch.fitted.converted)
mape(trainSet$month_Price, model.armax.garch.fitted.converted)
```

```{r}
rmse(testSet$month_Price, forecast.armax.garch.multi[1:length(testSet$Time)])
mae(testSet$month_Price, forecast.armax.garch.multi[1:length(testSet$Time)])
mape(testSet$month_Price, forecast.armax.garch.multi[1:length(testSet$Time)])
```


## 2.4 GAM Model
### 2.4.1 Fit Model
```{r}
trainSet$Time = as.Date(trainSet$Time)
trainSet$monthFac = as.factor(format(trainSet$Time, "%m"))
trainSet$Ndays = days_in_month(trainSet$Time)
trainSet$logdays = log(trainSet$Ndays)

gam_model <- gam(diff_log_price ~ s(as.numeric(Time), k=12) + s(Avg_Temp) + s(exchange_rate) +
                   s(monthFac, bs = "re") + sinpi(yday(Time) / 182.625) + 
                   cospi(yday(Time) / 182.625) + sinpi(yday(Time) / 91.3125) + 
                   cospi(yday(Time) / 91.3125) + offset(logdays),
                   data = trainSet, method = "ML", family = gaussian())

gam_model2 <- gam(diff_log_price ~ s(as.numeric(Time), k=100) + s(Avg_Temp) + 
                   s(log(exchange_rate)) + s(monthFac, bs = "re") + sinpi(yday(Time) / 182.625) + 
                   cospi(yday(Time) / 182.625) + sinpi(yday(Time) / 91.3125) + 
                   cospi(yday(Time) / 91.3125) + offset(logdays),
                   data = trainSet, method = "REML")

gam_model3 <- gam(diff_log_price ~ s(as.numeric(Time), k=100) + s(Avg_Temp) + 
                   s(log(exchange_rate)) + s(monthFac, bs = "re") + 
                    s(yday(Time), bs = "cc", k = 10) + offset(logdays),
                   data = trainSet, method = "REML")
```

```{r}
gam.check(gam_model)
gam.check(gam_model2)
gam.check(gam_model3)
```
```{r}
summary(gam_model)
summary(gam_model2)
summary(gam_model3)
```

### 2.4.2 Forecast and Plot
```{r}
testSet$Time = as.Date(testSet$Time)
testSet$monthFac = as.factor(format(testSet$Time, "%m"))
# trainSet$month_num = as.numeric(trainSet$monthFac)
# trainSet$timeNumeric = as.numeric(trainSet$date)
testSet$Ndays = days_in_month(testSet$Time)
testSet$logdays = log(testSet$Ndays)

testSet2 <- testSet
testSet2$log_exchange_rate <- log(testSet2$exchange_rate)
```

```{r}
# gam1
testSet$pred_log <- predict(gam_model, newdata = testSet)
testSet$pred_log_price <- last_log_price + cumsum(testSet$pred_log)
testSet$pred_price <- exp(testSet$pred_log_price)

# gam2
testSet2$pred_log2 <- predict(gam_model2, newdata = testSet2)
testSet2$pred_log_price2 <- last_log_price + cumsum(testSet2$pred_log2)
testSet2$pred_price2 <- exp(testSet2$pred_log_price2)

# gam3
testSet2$pred_log3 <- predict(gam_model3, newdata = testSet2)
testSet2$pred_log_price3 <- last_log_price + cumsum(testSet2$pred_log3)
testSet2$pred_price3 <- exp(testSet2$pred_log_price3)
```

```{r}
ggplot(testSet, aes(x = Time)) +
  geom_line(aes(y = month_Price, color = "Actual")) +
  geom_line(aes(y = pred_price, color = "Predicted")) +
  labs(title = "GAM Model 1: Forecast vs Actual Prices", 
       x = "Time", y = "Price") +
  scale_color_manual(values = c("Actual" = "black", "Predicted" = "red")) +
  theme_minimal()
```

```{r}
p4 <- ggplot(testSet2, aes(x = Time)) +
  geom_line(aes(y = month_Price, color = "Actual")) +
  geom_line(aes(y = pred_price2, color = "Prediction")) +
  labs(title = "GAM: Predicted vs Actual Prices", 
       x = "Time", y = "Price") +
  scale_color_manual(values = c("Actual" = "black", "Prediction" = "red")) +
  theme_minimal()
p4
```

```{r}
ggplot(testSet2, aes(x = Time)) +
  geom_line(aes(y = month_Price, color = "Actual")) +
  geom_line(aes(y = pred_price3, color = "Predicted")) +
  labs(title = "GAM Model 3: Actual vs. Predicted Price", 
       x = "Time", y = "Price") +
  scale_color_manual(values = c("Actual" = "black", "Predicted" = "red")) +
  theme_minimal()
```

```{r}
anova(gam_model, gam_model2, gam_model3)
```

```{r}
# RMSE
sqrt(mean((testSet$month_Price - testSet$pred_price)^2))
sqrt(mean((testSet2$month_Price - testSet2$pred_price2)^2))
sqrt(mean((testSet2$month_Price - testSet2$pred_price3)^2))

# MAE
mean(abs(testSet$month_Price - testSet$pred_price))
mean(abs(testSet2$month_Price - testSet2$pred_price2))
mean(abs(testSet2$month_Price - testSet2$pred_price3))

# MAPE
mean(abs((testSet$month_Price - testSet$pred_price) / testSet$month_Price)) * 100
mean(abs((testSet2$month_Price - testSet2$pred_price2) / testSet2$month_Price)) * 100
mean(abs((testSet2$month_Price - testSet2$pred_price3) / testSet2$month_Price)) * 100
```


## 2.5 Walk-Forward Validation with XGBoost Model
### 2.5.1 Fit and Forecast
```{r}
ntest <- nrow(data) - cutoff
predictions <- c()
actuals <- c()
dates <- c()
```

```{r}
data$monthFac <- as.factor(format(data$Time, "%m"))
data$Time <- as.numeric(as.Date(data$Time))
data$monthFac <- as.numeric(data$monthFac)
data$log_exchange_rate <- log(data$exchange_rate)
features <- c("monthFac", "Time", "Avg_Temp", "log_exchange_rate")
for (i in 1:ntest) {
  train_data <- data[1:(cutoff + i - 1), ]
  test_data <- data[(cutoff + i), ]
  x_train <- train_data %>% select(all_of(features))
  y_train <- train_data$log_price
  x_test <- test_data %>% select(all_of(features))
  dtrain <- xgb.DMatrix(data = as.matrix(x_train), label = y_train)
  dtest <- xgb.DMatrix(data = as.matrix(x_test))
  xgb_model <- xgboost(data = dtrain, nrounds = 100, objective = "reg:squarederror", verbose = 0)
  pred_log <- predict(xgb_model, dtest)
  pred_price <- exp(pred_log)
  predictions <- c(predictions, pred_price)
  actuals <- c(actuals, exp(test_data$log_price))
  dates <- c(dates, test_data$Time)
}

xgb_walk_df <- tibble(Time = as.Date(dates),
                      Actual = actuals,
                      Predicted = predictions)
```

```{r}
p5 <- ggplot(xgb_walk_df, aes(x = Time)) + geom_line(aes(y = Actual, color = "Actual")) + 
  geom_line(aes(y = Predicted, color = "Prediction")) +
  labs(title = "XGBoost: Predicted vs Actual Prices", x = "Time", y = "Price") + 
  scale_color_manual(values = c("Actual" = "black", "Prediction" = "red")) +
  theme_minimal()
p5
```

```{r}
# RMSE
sqrt(mean((xgb_walk_df$Actual - xgb_walk_df$Predicted)^2))

# MAE
mean(abs(xgb_walk_df$Actual - xgb_walk_df$Predicted))

# MAPE
mean(abs((xgb_walk_df$Actual - xgb_walk_df$Predicted) / xgb_walk_df$Actual)) * 100
```

## 2.6 Model Selection

```{r}
p1 <- p1 + theme(text = element_text(size = 8))  
p2 <- p2 + theme(text = element_text(size = 8))
p3 <- p3 + theme(text = element_text(size = 8))
p4 <- p4 + theme(text = element_text(size = 8))
p5 <- p5 + theme(text = element_text(size = 8))

p.combined <- grid.arrange(p1, p4, p2, p5, p3, 
             ncol = 2, 
             widths = c(2, 2),
             heights = c(1.2, 1.2, 1.2))
ggsave("plot_all_models.png", p.combined, dpi = 300)
```

# 3. Prediction
```{r}
# Predict 12 months
future_months <- 12

last_row <- data[nrow(data), ]
future_predictions <- c()
future_dates <- c()

for (i in 1:future_months) {
  future_time <- last_row$Time + (i * 30)
  future_monthFac <- as.numeric(format(as.Date(future_time, origin = "1970-01-01"), "%m"))

  future_data <- last_row
  future_data$Time <- future_time
  future_data$monthFac <- future_monthFac

  x_future <- future_data %>% select(all_of(features))
  dfuture <- xgb.DMatrix(data = as.matrix(x_future))

  pred_log_future <- predict(xgb_model, dfuture)
  pred_price_future <- exp(pred_log_future)

  future_predictions <- c(future_predictions, pred_price_future)
  future_dates <- c(future_dates, as.Date(future_time, origin = "1970-01-01"))

  last_row$log_price <- pred_log_future
}

future_xgb_df <- tibble(Time = as.Date(future_dates),
                        Predicted = future_predictions)

combined_df <- bind_rows(xgb_walk_df, future_xgb_df)
```

```{r}
forecast_start <- min(future_xgb_df$Time)
forecast_end <- max(future_xgb_df$Time)

ggplot(combined_df, aes(x = Time)) + 
  geom_rect(aes(xmin = forecast_start, xmax = forecast_end, ymin = -Inf, ymax = Inf), 
            fill = "lightgray", alpha = 0.3) + 
  geom_line(aes(y = Actual, color = "Actual"), na.rm = TRUE) +
  geom_line(aes(y = Predicted, color = "Predicted"), linetype = "dashed") +
  labs(title = "Future Predictions (12 months) using XGBoost Walk-Forward Forecast",
       x = "Time", y = "Price") +
  scale_color_manual(values = c("Actual" = "black", "Predicted" = "red")) +
  theme_minimal()
```

```{r}
summary(combined_df)
```

